The Calculus of Linear Constructions

### 2021-10-09 writing/draft1.tex
coq

% document body
\begin{document}
  \maketitle 
  \begin{abstract}
    The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends CC with a hierarchy of linear universes that precisely controls the introduction and elimination of its term level inhabitants. We study the meta-theory of CLC, proving that it is a sound logical framework for reasoning about resource. CLC is highly backwards compatible with CC, allowing CLC to enjoy the decades of CC research. We have formalized all major results in the Coq Proof Assistant. We extend CLC with linear inductive types and show that CLC as a programming language enables the manipulation of mutable data structures in a principled way.
  \end{abstract}

@@

  However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Intuitively, a mathematical theorem can be applied an unrestricted number of times. Comparatively, the usage of resources is more limited. For example, if we encode Girard's classical example \cite{girard95} of purchasing cigarettes literally into CC as a function of type:

  \begin{equation*}
    Money \rightarrow Camels \times Marlboro
  \end{equation*}

  \noindent The customer will still maintain full ownership of their money after paying the vendor. Unless the vendor is exceedingly generous, we are faced with the crime of counterfeiting. Users of proof assistants based on CC often need to embed external logics into CC \cite{bedrock} to provide additional reasoning principles for dealing with resource. The design and embedding of these logics is a difficult problem in its own right, requiring additional proofs to justify its soundness. We propose an alternative solution: extend CC with linear types.

  Linear Logic is a substructural logic introduced by Girard in his seminal work \cite{girard}. Girard notice that the Weakening and Contraction rules of Classical Logic when restricted carefully, gives rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} first notice that an analogous restriction to variable usage in simple type theory leads to a linear type theory, where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications. The success of integrating Linear Logic with simple type theory exposes a tantalizing new frontier of integrating linearity with richer type theories.

@@

  We propose a new linear dependent type system - The Calculus of Linear Constructions (CLC). Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``minimally invasive'' approach to extending CC with linear types. This allows for a straightforward embedding of CLC into CC, and lifting of CC into CLC, endowing CLC with the fruits of decades of CC research. We accomplish this by adding extra universes $L$ of linear types with cumulativity parallel to the universes $U$ of non-linear types. Universe information is propagated by an indexed typing judgment down to the term level, controlling the usage of Weakening and Contraction rules. We further extend CLC with inductive types, showing that as a programming language it can manipulate mutable data structures in a principled way. We have formalized all major results in Coq, and implemented a prototype in OCaml.

>>>>

% document body
\begin{document}
  \maketitle 
  \begin{abstract}
    The Calculus of Linear Constructions (CLC) is an extension of the Calculus of Constructions (CC) with linear types. Specifically, CLC extends CC with a hierarchy of linear universes that precisely controls the weakening and contraction of its term level inhabitants. We study the meta-theory of CLC, proving that it is a sound logical framework for reasoning about resource. CLC is highly backwards compatible with CC, allowing CLC to enjoy the decades of CC research. We have formalized and proven correct all major results of the core calculus in the Coq Proof Assistant. We extend CLC with linear inductive types and show that CLC as a programming language enables the manipulation of mutable data structures in a principled way.
  \end{abstract}

@@

  However, due to its origins as a logical framework for constructive mathematics, it is quite difficult for CC to encode and reason about resources. Intuitively, a mathematical theorem can be applied an unrestricted number of times. Comparatively, the usage of resources is more limited. For example, if we encode Girard's classical example \cite{girard95} of purchasing cigarettes literally into CC as a function of type:
  \begin{equation*}
    Money \rightarrow Camels + Marlboro
  \end{equation*}
  The customer will still maintain full ownership of their money after paying the vendor. Unless the vendor is exceedingly generous, we are faced with the crime of counterfeiting. Users of proof assistants based on CC often need to embed external logics into CC \cite{bedrock} to provide additional reasoning principles for dealing with resource. The design and embedding of these logics is a difficult problem in its own right, requiring additional proofs to justify its soundness. We propose an alternative solution: extend CC with linear types.

  Linear Logic is a substructural logic introduced by Girard in his seminal work \cite{girard}. Girard notice that the weakening and contraction rules of Classical Logic when restricted carefully, gives rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} first notice that an analogous restriction to variable usage in simple type theory leads to a linear type theory, where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications. The success of integrating Linear Logic with simple type theory exposes a tantalizing new frontier of integrating linearity with richer type theories.

@@

  We propose a new linear dependent type system - The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add extra universes $L$ of linear types with cumulativity parallel to the universes $U$ of non-linear types. Universe information is propagated by an indexed typing judgment down to the term level, controlling the usage of weakening and contraction rules. This ultimately results in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encodes the usage of resource. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$, and a linear arrow $\multimap_L$ as follows.
  \begin{equation*}
    Money\ 5 \multimap_L Camels + Marlboro
  \end{equation*}
  This new specification for transaction states that it requires a payment of 5 units of money, the customer is relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

  Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``minimally invasive'' approach to extending CC with linear types, akin to Mazurak et al.'s work on System F \cite{mazurak}. This allows for a straightforward embedding of CLC into CC, and lifting of CC into CLC, endowing CLC with the fruits of decades of CC research. We further extend CLC with inductive types, showing that as a programming language it can manipulate mutable data structures in a principled way. We have formalized all major results in Coq, and implemented a prototype in OCaml.
